Can We Crunch the Social Graph on the Cheap?

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex

%todo format the UW names horizontally
\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Dan Grossman, Mark Oskin,
Carl Ebeling, Luis Ceze}{University of Washington}{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}
\authorinfo{Simon Kahan}{Pacific Northwest National Lab}{skahan@cs.washington.edu}
\authorinfo{Preston Briggs}{}{preston.briggs@gmail.com}

\maketitle
\begin{abstract}
% Techniques from graph analytics have found important applications in
% areas such as social networks and bioinformatics. These irregular
% problems present a challenge for parallel machines: obtaining
% performance is not straightforward. Multithreading works well for
% such problems, but the only large-scale implementation of the technique
% is the Cray XMT, an expensive, non-commodity machine.

@@

%\todo{Begin eval insert}
%\input{evaluation}
%\todo{End eval insert}

\section{Related work}
\label{sec:related}
The concept of using independent parallel contexts to tolerate latency
is not novel.  The inspiration for our project is of course the Cray
XMT \cite{feo-xmt}, which implemented a hardware version of our
programming model (in what would today be considered exceptionally
slow silicon.)  The XMT can be seen as simultaneous multithreading
\cite{tullsen-smt} writ large; the technique and motivation is
essentially the same, but the XMT allows hundreds of contexts (modern
Intel SMT implementations are generally limited to two.) The Alewife
system \cite{agarwal-alewife} also switches contexts (in hardware) to
hide latency on remote-node accesses.  Cyclops \cite{almasi-cyclops}
allows hundreds of independently executing contexts to share FPUs,
caches, and paths to memory, but does not context switch or share
basic execution hardware.

Modern programmable GPUs, such as nVidia's CUDA programming model, can also be interpreted in using an XMT-like
model under the hood;
% Do I need to include a citation here?
while they generate large numbers of available execution contexts via
a (semi-)explicit SIMD model, those SIMD instructions keep functional
units satisfied by overlapping execution with the loads required for
other contexts.

Similar techniques have also been used in software.  Mowry proposed a system \cite{mowry-scm} of \emph{software-controlled
  multithreading}.  As opposed to our system of explicitly marked
remote references, SCM assumes a high hit rate and traps to an
interrupt routine which performed a context switch upon a cache miss.  The 2000-era processors had
short enough latencies that using more than two contexts generally
produced slowdown.

Partitioned global address space models, such as that used in Unified Parallel C,
% definitely need a citation here
 share our approach of exposing all data (even remote data)
 transparently to any threads and have largely solved the problems of
 how to efficiently translate those remote reads into message
 passing.  However, their programming models emphasize locality more
 than our work does; while a program can examine any vertex in a UPC
 program at any time, it probably shouldn't (whereas our kernels rely on our
 latency hiding technique to make common remote references fast.

The value of the information in extremely large and sparse graphs is
rapidly growing.  The GraphCT toolkit \cite{ediger-graphct} used an XMT to perform
sophisticated analysis on data from Twitter, identifying important
actors and their networks of influence, and it can scale to the full
size of graphs such as Facebook's friend network.

>>>>

%todo format the UW names horizontally
\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Preston Briggs,
  Dan Grossman, Mark Oskin, Carl Ebeling, Simon Kahan, Luis
  Ceze}{University of Washington}{\{nelson, bdmyers, ahh, preston,
  djg, oskin, ebeling, skahan, luisceze\}@cs.washington.edu}

\maketitle

\begin{abstract}
% Techniques from graph analytics have found important applications in
% areas such as social networks and bioinformatics. These irregular
% problems present a challenge for parallel machines: obtaining
% performance is not straightforward. Multithreading works well for
% such problems, but the only large-scale implementation of the technique
% is the Cray XMT, an expensive, non-commodity machine.

@@

\section{Related work}
\label{sec:related}

Much work exists on using multithreading to tolerate latency. Hardware
implementations include the Tera MTA~\cite{tera} and Cray
XMT~\cite{feo-xmt}, Simultaneous Multithreading \cite{tullsen-smt},
MIT Alewife \cite{agarwal-alewife}, Cyclops \cite{almasi-cyclops}, and
even GPUs \cite{gpus}. Software-only approaches exist as well; the
Software Controlled Multithreading project \cite{mowry-scm}, QThreads
\cite{qthreads}, MAESTRO \cite{maestro}, and Charm \cite{charm} all
use variants of this idea.
%In contrast to all of these, our goal is to tolerate
%latencies across a multi-node system using mostly software.

Our goal of presenting a global view of distributed memory to the
programmer is shared by the PGAS community, and is used in languages
like Chapel \cite{chapel}, X10 \cite{X10}, and UPC \cite{upc}. They
obtain performance by minimizing references to remote nodes, whereas we
optimize for this case.

The idea of processing large graphs on distributed machines has been
exploredby projects such as Pregel \cite{pregel}, the Parallel Boost
Graph Library \cite{parallelbgl}, and CGMLib \cite{cgmlib}. Our goal
is to develop infrastructure to aid implementation of similar
libraries.
